---
title: 'Ch2: Small Worlds and Large Worlds'
author: "Samir Gadkari"
date: "2/21/2021"
output: html_document
---

```{r}
library(tidyverse)
```


### 2.4.3 Grid Approximation for the globe tossing experiment

```{r}
p_grid <- seq(0, 1, length.out = 20) # probability grid

prior <- function(type) {
  if (type == 1L) {
    
    rep(1, 20)                   # First prior - uniform distribution
  } else if (type == 2L) {
    
    ifelse(p_grid < 0.5, 0, 1)   # Second prior - step distribution
  } else {
    
  exp(-5 * abs(p_grid - 0.5))    # Third prior - gaussian distribution
  }
}

# dbinom gives the density of the binomial distribution, given:
#   the number of successes
#   number of trials
#   the probability of success
# Here we're calculating the likelihood at each point on the grid.
likelihood <- dbinom(6,         # number of successes
                     size = 9,  # number of trials
                     prob = p_grid) # probability of success per trial
# If you run this:
# dbinom(3, 3, prob = p_grid), you're saying that we expect all trails
# to be successful. So dbinom gives the 20th plausibility the largest value,
# (all 20 successes), and subsequent lower plausibilities 
# lower and lower values.
# dbinom(3, 4, prob = p_grid), you're saying that we expect about 3 out of
# 4 trials to be successful, so dbinom gives a higher probability to 
# 15 successes out of 20, and lower probabilities for higher and lower
# successes.
# 
# Likelihood has nothing to do with the prior !!!!!

plot_posterior <- function(type) {
  # Un-standardized posterior distribution
  unstd.posterior <- prior(type) * likelihood
  
  # Standardize the posterior distribution
  posterior <- unstd.posterior / sum(unstd.posterior)
  
  plot(p_grid, posterior, 
       type = "b",   # 'p' : points, "l" : lines, "b" : both
       xlab = "probability of water",
       ylab = "posterior probability"
  )
  mtext("20 points")
}
```
```{r}
plot_prior <- function(p_grid, prior_type) {
  plot(p_grid, prior(prior_type), type = "b")
  mtext(paste("prior_type =", prior_type))
}
plot_likelihood <- function() {
  plot(p_grid, likelihood, type = "b")
}
```

```{r}
plot_likelihood()
plot_prior(p_grid, prior_type = 1)
plot_posterior(1)
plot_prior(p_grid, prior_type = 2)
plot_posterior(2)
plot_prior(p_grid, prior_type = 3)
plot_posterior(3)
```

### 2.4.4 Quadratic approximation for the globe tossing experiment

To compute the quadratic approximation to the globe tossing posterior:
```{r}
library(rethinking)

globe.qa <- quap(alist(
                    W ~ dbinom(W + L, p), # binomial likelihood
                    p ~ dunif(0, 1)       # uniform prior
                  ),
                 data = list(W = 6, L = 3)
                )
precis(globe.qa)   # display summary of quadratic approximation
```

Posterior mean value = p = 0.67
Curvature of the quadratic approximation = sd = 0.16
89 percentile interval = [0.42, 0.92]

Plotting the quadratic approximation against the actual posterior:
```{r}
W <- 6
L <- 3

curve(dbeta(x, W + 1, L + 1), from = 0, to = 1) # x goes from 0 through 1,
                                                # because of curve's params
                                                # from and to.
curve(dnorm(x, 0.67, 0.16), lty = 2, add = TRUE) # Since add is TRUE,
                                                 # x takes the same values
                                                 # as the earlier plot
```
Note: The dnorm output is positive at 1.0, which is impossible.
A positive value of 1 would mean only water is possible.
This can't be true, since our data saw land.
This is why many scientists require number of samples > 30.
But some models are only correct with thousands of samples, so this
heuristic is not valid. Instead of using quadratic approximation,
you could use MCMC (Markov Chain Monte Carlo), or a whole host of
other techniques.

The quadratic approximation with lots of data and a uniform prior =
the Maximum Likelihood Estimate (MLE) and it's Standard Error. Thus, we can
re-interpret a lot of non-Bayesian model in Bayesian terms. But MLE 
has some drawbacks that the quadratic approximation shares. This is why
we turn to MCMC for the second half of the book.

### 2.4.5 Markov Chain Monte Carlo

MCMC (Markov Chain Monte Carlo) approximation to the globe tossing 
posterior:
```{r}
n_samples <- 1000

p <- rep(NA, n_samples)
p[1] <- 0.5

W <- 6
L <- 3

for (i in 2:n_samples) {
  p_new <- rnorm(1,         # get 1 sample 
                 p[i - 1],  # with this mean
                 0.1)       # and this sd
  if (p_new < 0) p_new <- abs(p_new) # Keep p_new between 0, 1.
  if (p_new > 1) p_new <- 2 - p_new
  
  q0 <- dbinom(W, W + L, p[i - 1])
  q1 <- dbinom(W, W + L, p_new)
  
  p[i] <- ifelse(runif(1) < q1 / q0, p_new, p[i - 1])
}
```

The values from p are samples from the posterior. To compare with the
analytic posterior:
```{r}
dens(p, xlim = c(0, 1)) # dens is just a function to plot densities.
                        # It plots p here.

# First argument of curve() is the expression to draw.
curve(dbeta(x, W + 1, L + 1), # Density of the beta distribution with
                              # shape1 = W + 1, shape2 = L + 1.
      lty = 2,    # lty (linetype) = 2 for a dashed line
      add = TRUE) # If add = TRUE, add to an already existing plot
```

This is the Metropolis algorithm, and is explained in Chapter 9.

## 2.6 Practice

2E1. Which of the expressions below correspond to the statement: the probability of rain on Monday?

Pr(rain | Monday)

Since:
Pr(rain, Monday) = Pr(rain | Monday) Pr(Monday), and
Pr(rain, Monday) = Pr(Monday | rain) Pr(rain), this is also
equivalent to:
Pr(rain, Monday) / Pr(Monday)

2E2. Which of the following statements corresponds to the expression: Pr(Monday|rain)?

The probability that it is Monday, given that it is raining.

2E3. Which of the expressions below correspond to the statement: the probability that it is Monday, given that it is raining?

Pr(Monday | rain)

Since Pr(Monday | rain) P(rain) = Pr(rain | Monday) Pr(Monday), we can say:
Pr(Monday | rain) = Pr(rain | Monday) Pr(Monday) / Pr(rain)

2M1. Recall the globe tossing model from the chapter. Compute and plot the grid approximate posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for p.
(1) W, W, W
(2) W, W, W, L
(3) L, W, W, L, W, W, W”

Let's go through the whole procedure again, but in a different way:
```{r}
p_grid <- seq(0, 1, length.out = 20)
prior <- rep(1, 20)
trials <- tibble(water = c(3, 3, 5), land = c(0, 1, 2))

grid_approx <- function(p_grid, prior, trials) {
  for (i in 1:nrow(trials)) {
    likelihood <- dbinom(trials["water"][[1]][[i]], 
                         trials["water"][[1]][[i]] + trials["land"][[1]][[i]],
                         p_grid)
    unstd_posterior <- prior * likelihood
    posterior <- unstd_posterior / sum(unstd_posterior)
    plot(p_grid, posterior, type = "b")
  }
}

grid_approx(p_grid, prior, trials)
```

2M2. Now assume a prior for p that is equal to zero when p < 0.5 and is a positive constant when p ≥ 0.5. Again compute and plot the grid approximate posterior distribution for each of the sets of observations in the problem just above.

```{r}
prior <- if_else(p_grid < 0.5, 0, 1)
grid_approx(p_grid, prior, trials)
```

2M3. Suppose there are two globes, one for Earth and one for Mars. The Earth globe is 70% covered in water. The Mars globe is 100% land. Further suppose that one of these globes—you don’t know which—was tossed in the air and produced a “land” observation. Assume that each globe was equally likely to be tossed. Show that the posterior probability that the globe was the Earth, conditional on seeing “land” (Pr(Earth|land)), is 0.23.

Pr(Earth)        = 0.5
Pr(Mars)         = 0.5
Pr(land | Earth) = 0.3  (given Earth - that means we don't look at
                         the probability of getting to Earth is 0.5)
Pr(land | Mars)  = 1.0  (given Mars  - that means we don't look at
                         the probability of getting to Mars is 0.5)

Pr(data) = Integral(Pr(data | prior) * Pr(pior))
Pr(land) = Pr(land | Earth) * Pr(Earth) +  Pr(land | Mars) * Pr(Mars)
Pr(land) =         0.3      *    0.5    +         1.0      *    1.0
         =                0.15          +                 0.5
         =                             0.65
Pr(Earth | land) = Pr(land | Earth) * Pr(Earth) / Pr(land)
                 =         0.3      *     0.5   /   0.65
                 =                     0.23

2M4. Suppose you have a deck with only three cards. Each card has two sides, and each side is either black or white. One card has two black sides. The second card has one black and one white side. The third card has two white sides. Now suppose all three cards are placed in a bag and shuffled. Someone reaches into the bag and pulls out a card and places it flat on a table. A black side is shown facing up, but you don’t know the color of the side facing down. Show that the probability that the other side is also black is 2/3. Use the counting method (Section 2 of the chapter) to approach this problem. This means counting up the ways that each card could produce the observed data (a black side facing up on the table).

3 cards in a deck
Each card side is either black or white.
One card has 2 black sides
One card has 1 black and 1 white
One card has 2 white sides

Observed = one card with black side facing up

Show that Pr(other side is also black) = 2/3

Pr(data) = Integral(Pr(black | card n) * Pr(card n))
         =  1.0 * 0.33 + 0.5 * 0.33 + 0 * 0.33
         = 0.495
         
Pr(both sides black | black side up) = 
  Pr(black side up | both sides black) * Pr(both sides black) /
    Pr(black side up)
  = 1.0 * 0.33 / 0.495
  = 0.66
  = 2 / 3
  
2M5. Now suppose there are four cards: B/B, B/W, W/W, and another B/B. Again suppose a card is drawn from the bag and a black side appears face up. Again calculate the probability that the other side is black.

Observed = one card with black side facing up

Pr(other side is also black) = ?

Pr(data) = 1.0 * 0.25 + 0.5 * 0.25 + 0 * 0.25 + 1.0 * 0.25
         = 0.625
Pr(both sides black | black side up) = 
  Pr(black side up | both sides black) * Pr(both sides black) /
    Pr(black side up)
  = 1.0 * 0.5 / 0.625
  = 0.8
  
2M6. Imagine that black ink is heavy, and so cards with black sides are heavier than cards with white sides. As a result, it’s less likely that a card with black sides is pulled from the bag. So again assume there are three cards: B/B, B/W, and W/W. After experimenting a number of times, you conclude that for every way to pull the B/B card from the bag, there are 2 ways to pull the B/W card and 3 ways to pull the W/W card. Again suppose that a card is pulled and a black side appears face up. Show that the probability the other side is black is now 0.5. Use the counting method, as before.

Pr(BB) = 1 / 6
Pr(BW) = 2 / 6
Pr(WW) = 3 / 6

Pr(data) = Pr(black side up) = Integral(Pr(black | card n) * Pr(card n))
         = 1 * 1 / 6 + 0.5 * 2 / 6 + 0 * 3 / 6
```{r}
1 * 1 / 6 + 0.5 * 2 / 6 + 0 * 3 / 6
```

Pr(both sides black | black side up) = 
  Pr(black side up | both sides black) Pr(both sides black) /
    Pr(black side up)
  = 1.0 * (1 / 6) / 0.33
```{r}
1.0 * (1 / 6) / 0.33
```

2M7. Assume again the original card problem, with a single card showing a black side face up. Before looking at the other side, we draw another card from the bag and lay it face up on the table. The face that is shown on the new card is white. Show that the probability that the first card, the one showing a black side, has black on its other side is now 0.75. Use the counting method, if you can. Hint: Treat this like the sequence of globe tosses, counting all the ways to see each observation, for each possible first card.

Pr(BB) = 1 / 6
Pr(BW) = 2 / 6
Pr(WW) = 3 / 6
                             BB        BW
First card top side black = 1/6 * 1 + 2/6 * 0.5 = 2/6 = 1/3

                            WW         BW      
Second card top side W    = 1 * 0.5 + 0.5 * 0.5 = 0.5 + 0.25 = 0.75

First card top side black and Second card top side W = 0.333 * 0.75 = 0.739
First card other side black = 0.739 * (1 + )
