---
title: 'Ch2: Small Worlds and Large Worlds'
author: "Samir Gadkari"
date: "2/21/2021"
output: html_document
---

```{r}
library(tidyverse)
```


### 2.4.3 Grid Approximation

```{r}
p_grid <- seq(0, 1, length.out = 20) # probability grid

prior <- function(type) {
  if (type == 1L) {
    
    rep(1, 20)                   # First prior - uniform distribution
  } else if (type == 2L) {
    
    ifelse(p_grid < 0.5, 0, 1)   # Second prior - step distribution
  } else {
    
  exp(-5 * abs(p_grid - 0.5))    # Third prior - gaussian distribution
  }
}

# dbinom gives the density of the binomial distribution, given:
#   the number of successes
#   number of trials
#   the probability of success
# Here we're calculating the likelihood at each point on the grid.
likelihood <- dbinom(6,         # number of successes
                     size = 9,  # number of trials
                     prob = p_grid) # probability of success per trial
# If you run this:
# dbinom(3, 3, prob = p_grid), you're saying that we expect all trails
# to be successful. So dbinom gives the 20th plausability the largest value,
# (all 20 successes), and subsequent lower plausabilities 
# lower and lower values.
# dbinom(3, 4, prob = p_grid), you're saying that we expect about 3 out of
# 4 trials to be successful, so dbinom gives a higher probability to 
# 15 successes out of 20, and lower probabilities for higher and lower
# successes.
# 
# Likelihood has nothing to do with the prior !!!!!

plot_posterior <- function(type) {
  # Unstandardized posterior distribution
  unstd.posterior <- prior(type) * likelihood
  
  # Standardize the posterior distribution
  posterior <- unstd.posterior / sum(unstd.posterior)
  
  plot(p_grid, posterior, 
       type = "b",   # 'p' : points, "l" : lines, "b" : both
       xlab = "probability of water",
       ylab = "posterior probability"
  )
  mtext("20 points")
}
```
```{r}
plot_prior <- function(p_grid, prior_type) {
  plot(p_grid, prior(prior_type), type = "b")
  mtext(paste("prior_type =", prior_type))
}
plot_likelihood <- function() {
  plot(p_grid, likelihood, type = "b")
}
```

```{r}
plot_likelihood()
plot_prior(p_grid, prior_type = 1)
plot_posterior(1)
plot_prior(p_grid, prior_type = 2)
plot_posterior(2)
plot_prior(p_grid, prior_type = 3)
plot_posterior(3)
```

### 2.4.4 Quadratic approximation

To compute the quadratic approximation to the globe tossing posterior:
```{r}
library(rethinking)

globe.qa <- quap(alist(
                    W ~ dbinom(W + L, p), # binomial likelihood
                    p ~ dunif(0, 1)       # uniform prior
                  ),
                 data = list(W = 6, L = 3)
                )
precis(globe.qa)   # display summary of quadratic approximation
```

Posterior mean value = p = 0.67
Curvature of the quadratic approximation = sd = 0.16
89 percentile interval = [0.42, 0.92]

Plotting the quadratic approximation against the actual posterior:
```{r}
W <- 6
L <- 3

curve(dbeta(x, W + 1, L + 1), from = 0, to = 1)
curve(dnorm(x, 0.67, 0.16), lty = 2, add = TRUE)
```
Note: The dnorm output is positive at 1.0, which is impossible.
A positive value of 1 would mean only water is possible.
This can't be true, since our data saw land.
This is why many scientists require number of samples > 30.
But some models are only correct with thousands of samples, so this
heuristic is not valid. Instead of using quadratic approximation,
you could use MCMC (Markov Chain Monte Carlo), or a whole host of
other techniques.

The quadratic approximation with lots of data and a uniform prior =
the Maximum Likelihood Estimate (MLE) and it's Standard Error. Thus, we can
re-interpret a lot of non-Bayesian model in Bayesian terms. But MLE 
has some drawbacks that he quadratic approximation shares. This is why
we turn to MCMC for the second half of the book.

### 2.4.5 Markov Chain Monte Carlo

MCMC (Markov Chain Monte Carlo) approximation to the globe tossing 
posterior:
```{r}
n_samples <- 1000

p <- rep(NA, n_samples)
p[1] <- 0.5

W <- 6
L <- 3

for (i in 2:n_samples) {
  p_new <- rnorm(1,         # get 1 sample 
                 p[i - 1],  # with this mean
                 0.1)       # and this sd
  if (p_new < 0) p_new <- abs(p_new)
  if (p_new > 1) p_new <- 2 - p_new
  
  q0 <- dbinom(W, W + L, p[i - 1])
  q1 <- dbinom(W, W + L, p_new)
  
  p[i] <- ifelse(runif(1) < q1 / q0, p_new, p[i - 1])
}
```

The values from p are samples from the posterior. To compare with the
analytic posterior:
```{r}
dens(p, xlim = c(0, 1)) # dens is just a function to plot densities.
                        # It plots p here.

# First argument of curve() is the expression to draw.
curve(dbeta(x, W + 1, L + 1), # Density of the beta distribution with
                              # shape1 = W + 1, shape2 = L + 1.
      lty = 2,    # lty (linetype) = 2 for a dashed line
      add = TRUE) # If add = TRUE, add to an already existing plot
```

This is the Metropolis algorithm, and is explained in Chapter 9.

## 2.6 Practice

2E1. Which of the expressions below correspond to the statement: the probability of rain on Monday?

Pr(rain | Monday)

Since:
Pr(rain, Monday) = Pr(rain | Monday) Pr(Monday), and
Pr(rain, Monday) = Pr(Monday | rain) Pr(rain), this is also
equivalent to:
Pr(rain, Monday) / Pr(Monday)

2E2. Which of the following statements corresponds to the expression: Pr(Monday|rain)?

The probability that it is Monday, giving that it is raining.

2E3. Which of the expressions below correspond to the statement: the probability that it is Monday, given that it is raining?

Pr(Monday | rain)

Since Pr(Monday | rain) P(rain) = Pr(rain | Monday) Pr(Monday), we can say:
Pr(Monday | rain) = Pr(rain | Monday) Pr(Monday) / Pr(rain)

2M1. Recall the globe tossing model from the chapter. Compute and plot the grid approximate posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for p.
(1) W, W, W
(2) W, W, W, L
(3) L, W, W, L, W, W, W”

Let's go through the whole procedure again, but in a different way:
```{r}
p_grid <- seq(0, 1, length.out = 20)
prior <- rep(1, 20)
trials <- tibble(water = c(3, 3, 5), land = c(0, 1, 2))

grid_approx <- function(p_grid, prior, trials) {
  for (i in 1:nrow(trials)) {
    likelihood <- dbinom(trials["water"][[1]][[i]], 
                         trials["water"][[1]][[i]] + trials["land"][[1]][[i]],
                         p_grid)
    unstd_posterior <- prior * likelihood
    posterior <- unstd_posterior / sum(unstd_posterior)
    plot(p_grid, posterior, type = "b")
  }
}

grid_approx(p_grid, prior, trials)
```

2M2. Now assume a prior for p that is equal to zero when p < 0.5 and is a positive constant when p ≥ 0.5. Again compute and plot the grid approximate posterior distribution for each of the sets of observations in the problem just above.

```{r}
prior <- if_else(p_grid < 0.5, 0, 1)
grid_approx(p_grid, prior, trials)
```

